--- 
title: "TMA4250 Spatial Statistics Exercise 2, Spring 2019"
date: "`r format(Sys.time(), '%d.%m.%Y')`"
output:
  pdf_document: default
  bookdown::html_document2: default
  html_document:
    df_print: paged
subtitle: 'Group members: Henrik Syversveen Lie, Øyvind Auestad'
header-includes: \usepackage{float}
references:
- id: Omre
  title: Bayesian Spatial Inversion with Conjugate Prior Models
  author:
  - family: Omre
    given: Henning
  type: article-journal
  issued:
    year: 2019
    month: 1
---


```{r setup, include = FALSE}
library(bookdown)
library(formatR)
showsol <- FALSE
library(knitr)
opts_chunk$set(tidy.opts = list(width.cutoff = 68), tidy = TRUE, warning = FALSE, error = FALSE, message = FALSE, echo = TRUE, fig.pos = 'H')
```


```{r, echo = F, eval = T}
library(reshape2)
library(geoR)
library(ggplot2)
library(MASS)
library(cowplot)
library(fields)
library(akima)
library(spatial)
```

# Problem 1: Markov RF

This problem is based on observations of seismic data over a domain $\mathcal{D}\subset \mathbb{R}^2$. The objective is to identify the underlying {sand, shale} lithology distribution over $\mathcal{D}$, represented by $\{0, 1\}$ respectively. 

The observations are collected on a regular $(75 \times 75)$ grid $L_D$, and the seismic data are denoted $\{d(\mathbf{x}); \mathbf{x} \in L_D\};d(\mathbf{x})\in \mathbb{R}$, represented by the $n$-vector $\mathbf{d}$. We retrieve the observations from the R library MASS in the file _seismic.dat_. 

Moreover, observations of the lithology distribution {sand, shale} in a geologically comparable domain $\mathcal{D}_c \subset \mathbb{R}^2$ is available. THe lithology distribution is collected on a regular $(50\times 50)$ grid $L_{D_c}$, which has the same spacing as $L_D$ over $\mathcal{D}_c$. Here, we retrieve the observations from the same R library MASS in the file _complit.dat_.

We assume that the underlying lithology distribution can be represented by a Mosaic RF $\{l(\mathbf{x});\mathbf{x}\in L_D\};l(\mathbf{x})\in\{0,1\}$ represented by the $n$-vector $\mathbf{l}$.

## a)
