--- 
title: "TMA4250 Spatial Statistics Exercise 3, Spring 2019"
output:
  pdf_document:
    toc: no
    toc_depth: '2'
date: "`r format(Sys.time(), '%d.%m.%Y')`"
subtitle: 'Group members: Henrik Syversveen Lie, Øyvind Auestad'
header-includes: \usepackage{float}
references:
- id: Omre
  title: Bayesian Spatial Inversion with Conjugate Prior Models
  author:
  - family: Omre
    given: Henning
  type: article-journal
  issued:
    year: 2019
    month: 1
---


```{r setup, include = FALSE}
library(bookdown)
library(formatR)
showsol <- FALSE
library(knitr)
opts_chunk$set(tidy.opts = list(width.cutoff = 68), tidy = TRUE, warning = FALSE, error = FALSE, message = FALSE, echo = TRUE, fig.pos = 'H')
```


```{r, echo = F, eval = T}
library(reshape2)
library(geoR)
library(ggplot2)
library(MASS)
library(cowplot)
library(fields)
library(akima)
library(spatial)
```

# Problem 1: Markov RF

This problem is based on observations of seismic data over a domain $\mathcal{D}\subset \mathbb{R}^2$. The objective is to identify the underlying {sand, shale} lithology distribution over $\mathcal{D}$, represented by $\{0, 1\}$ respectively. 

The observations are collected on a regular $(75 \times 75)$ grid $L_D$, and the seismic data are denoted $\{d(\mathbf{x}); \mathbf{x} \in L_D\};d(\mathbf{x})\in \mathbb{R}$, represented by the $n$-vector $\mathbf{d}$. We retrieve the observations from the R library MASS in the file _seismic.dat_. 

Moreover, observations of the lithology distribution {sand, shale} in a geologically comparable domain $\mathcal{D}_c \subset \mathbb{R}^2$ is available. The lithology distribution is collected on a regular $(50\times 50)$ grid $L_{D_c}$, which has the same spacing as $L_D$ over $\mathcal{D}_c$. Here, we retrieve the observations from the same R library MASS in the file _complit.dat_.

We assume that the underlying lithology distribution can be represented by a Mosaic RF $\{l(\mathbf{x});\mathbf{x}\in L_D\};l(\mathbf{x})\in\{0,1\}$ represented by the $n$-vector $\mathbf{l}$.

## a)
The seismic data collection procedure defines the likelihood model:
$$[d_i|\mathbf{l}] = \begin{cases} 0.02+U_i, \text{ if } l_i = 0 - \text{ sand},\\ 0.08+U_i, \text{ if } l_i = 1 - \text{ shale}\end{cases}; i = 1,\dots,n,$$
with $U_i, i=1,\dots,n$ i.i.d. $N_1(0,0.06^2)$.


Since the $U_i$'s are all independent, $d_i$ depends only on $l_i$. Hence we get the following likelihood model
\begin{equation}
\begin{split}
[\mathbf{d} | \mathbf{l}] \sim p(\mathbf{d}|\mathbf{l}) &= \prod_{i = 1}^n p(d_i | \mathbf{l}) = \prod_{i = 1}^n p(d_i | l_i) = \prod_{i = 1}^n
\begin{cases}
N(d_i; 0.02, 0.06^2) \ &\textrm{if} \ l_i = 0 \\
N(d_i; 0.08, 0.06^2) \ &\textrm{if} \ l_i = 1
\end{cases} \\
&= \prod_{i : l_i = 0} N(d_i; 0.02, 0.06^2) \prod_{i:l_i = 1} N(d_i; 0.08, 0.06^2).
\end{split}
\label{eq:lik}
\end{equation}

We then display the observations as a map.
```{r, echo = F, eval = T, out.width = "50%", fig.align = "center", fig.cap = "\\label{fig:obs} Observed values of $\\mathbf{d}$ from the seismic data."}
dobs_df <- read.table("https://www.math.ntnu.no/emner/TMA4250/2017v/Exercise3/seismic.dat")

dlist = list(x = 1:75, y = 1:75, z = matrix(dobs_df[,1], nrow = 75, ncol = 75))
image.plot(dlist)
```

## b)

We now give $\mathbf{l}$ a uniform constant prior, i.e. $p(\mathbf{l}) = \textrm{const}$. We obtain the following posterior model
$$
[\mathbf{l} | \mathbf{d}] \sim p(\mathbf{l}|\mathbf{d}) = \frac{p(\mathbf{d} | \mathbf{l})}{\sum_{\mathbf{l}' \in L} p(\mathbf{d} | \mathbf{l}')},
$$
where the expression for the likelihood $p(\mathbf{d}|\mathbf{l})$ is given in equation \ref{eq:lik}. This gives
\begin{equation}
\begin{split}
p(\mathbf{l} | \mathbf{d}) &= \frac{ \prod_{i : l_i = 0} N(d_i; 0.02, 0.06^2) \prod_{i:l_i = 1} N(d_i; 0.08, 0.06^2) }{\sum_{\mathbf{k} \in \{0, 1\}^n} \prod_{i : k_i = 0} N(d_i; 0.02, 0.06^2) \prod_{i:k_i = 1} N(d_i; 0.08, 0.06^2) } \\
&= \frac{ \prod_{i : l_i = 0} \exp\big(-\frac{1}{2\cdot 0.06^2}(d_i - 0.02)^2\big) \prod_{i:l_i = 1} \exp\big(-\frac{1}{2\cdot 0.06^2}(d_i - 0.08)^2\big)}{\sum_{\mathbf{k} \in \{0, 1\}^n} \prod_{i : k_i = 0} \exp\big(-\frac{1}{2\cdot 0.06^2}(d_i - 0.02)^2\big) \prod_{i:k_i = 1} \exp\big(-\frac{1}{2\cdot 0.06^2}(d_i - 0.08)^2\big) }\\
&= \frac{ \exp\big(-\frac{1}{2\cdot 0.06^2}\big(\sum_{i:l_i = 0}(d_i - 0.02)^2 + \sum_{i:l_i = 1} (d_i - 0.08)^2\big) \big) }{\sum_{\mathbf{k} \in \{0, 1\}^n} \exp\big(-\frac{1}{2\cdot 0.06^2}\big(\sum_{i:k_i = 0}(d_i - 0.02)^2 + \sum_{i:k_i = 1} (d_i - 0.08)^2\big) \big)}.
\end{split}
\end{equation}

We want to simulate from the posterior Mosaic RF $\{l(\mathbf{x});\mathbf{x}\in L_D|\mathbf{d}\}$. Because we have an independent prior, each point in the Mosaic RF will be independent, and we get the following posterior distribution
$$[l_i|d_i]\sim p(l_i|d_i) = \frac{p(d_i|l_i)}{\sum_{l_i'\in L} p(d_i|l_i')} = \begin{cases} \frac{N(d_i;0.02,0.06^2)}{N(d_i;0.02,0.06^2)+N(d_i;0.08,0.06^2)}, \text{ if } l_i = 0,\\ \frac{N(d_i;0.08,0.06^2)}{N(d_i;0.02,0.06^2)+N(d_i;0.08,0.06^2)}, \text{ if } l_i = 1\end{cases}; i =1,\dots,n.$$
This means that we can simulate from the posterior distribution by using a single-site MCMC/Gibbs Simulation algorithm. We implement a version of Algorithm 9 from page 80 of @Omre.

# References
