--- 
title: "TMA4250 Spatial Statistics Exercise 3, Spring 2019"
output:
  pdf_document:
    toc: no
    toc_depth: '2'
date: "`r format(Sys.time(), '%d.%m.%Y')`"
subtitle: 'Group members: Henrik Syversveen Lie, Øyvind Auestad'
header-includes: \usepackage{float}
---


```{r setup, include = FALSE}
library(bookdown)
library(formatR)
showsol <- FALSE
library(knitr)
opts_chunk$set(tidy.opts = list(width.cutoff = 68), tidy = TRUE, warning = FALSE, error = FALSE, message = FALSE, echo = TRUE, fig.pos = 'H')
```


```{r, echo = F, eval = T}
library(reshape2)
library(geoR)
library(ggplot2)
library(MASS)
library(cowplot)
library(fields)
library(akima)
library(spatial)
```

# Problem 1: Markov RF

This problem is based on observations of seismic data over a domain $\mathcal{D}\subset \mathbb{R}^2$. The objective is to identify the underlying {sand, shale} lithology distribution over $\mathcal{D}$, represented by $\{0, 1\}$ respectively. 

The observations are collected on a regular $(75 \times 75)$ grid $L_D$, and the seismic data are denoted $\{d(\mathbf{x}); \mathbf{x} \in L_D\};d(\mathbf{x})\in \mathbb{R}$, represented by the $n$-vector $\mathbf{d}$. We retrieve the observations from the R library MASS in the file _seismic.dat_. 

Moreover, observations of the lithology distribution {sand, shale} in a geologically comparable domain $\mathcal{D}_c \subset \mathbb{R}^2$ is available. The lithology distribution is collected on a regular $(50\times 50)$ grid $L_{D_c}$, which has the same spacing as $L_D$ over $\mathcal{D}_c$. Here, we retrieve the observations from the same R library MASS in the file _complit.dat_.

We assume that the underlying lithology distribution can be represented by a Mosaic RF $\{l(\mathbf{x});\mathbf{x}\in L_D\};l(\mathbf{x})\in\{0,1\}$ represented by the $n$-vector $\mathbf{l}$.

## a)
The seismic data collection procedure defines the likelihood model:
$$[d_i|\mathbf{l}] = \begin{cases} 0.02+U_i, \text{ if } l_i = 0 - \text{ sand},\\ 0.08+U_i, \text{ if } l_i = 1 - \text{ shale}\end{cases}; i = 1,\dots,n,$$
with $U_i, i=1,\dots,n$ i.i.d. $N_1(0,0.06^2)$.


Since the $U_i$'s are all independent, $d_i$ depends only on $l_i$. Hence we get the following likelihood model
\begin{equation}
\begin{split}
[\mathbf{d} | \mathbf{l}] \sim p(\mathbf{d}|\mathbf{l}) &= \prod_{i = 1}^n p(d_i | \mathbf{l}) = \prod_{i = 1}^n p(d_i | l_i) = \prod_{i = 1}^n
\begin{cases}
N(d_i; 0.02, 0.06^2) \ &\textrm{if} \ l_i = 0 \\
N(d_i; 0.08, 0.06^2) \ &\textrm{if} \ l_i = 1
\end{cases} \\
&= \prod_{i : l_i = 0} N(d_i; 0.02, 0.06^2) \prod_{i:l_i = 1} N(d_i; 0.08, 0.06^2).
\end{split}
\label{eq:lik}
\end{equation}

We then display the observations as a map.
```{r, echo = F, eval = T, out.width = "50%", fig.align = "center", fig.cap = "\\label{fig:obs} Observed values of $\\mathbf{d}$ from the seismic data."}
obs <- read.table("https://www.math.ntnu.no/emner/TMA4250/2017v/Exercise3/seismic.dat")[,1]


dlist = list(x = 1:75, y = 1:75, z = matrix(obs, nrow = 75, ncol = 75))
image.plot(dlist)
```

## b)

We now give $\mathbf{l}$ a uniform constant prior, i.e. $p(\mathbf{l}) = \textrm{const}$. We obtain the following posterior model
$$
[\mathbf{l} | \mathbf{d}] \sim p(\mathbf{l}|\mathbf{d}) = \frac{p(\mathbf{d} | \mathbf{l})}{\sum_{\mathbf{l}' \in L} p(\mathbf{d} | \mathbf{l}')},
$$
where the expression for the likelihood $p(\mathbf{d}|\mathbf{l})$ is given in equation \ref{eq:lik}. This gives
\begin{equation}
\begin{split}
p(\mathbf{l} | \mathbf{d}) &= \frac{ \prod_{i : l_i = 0} N(d_i; 0.02, 0.06^2) \prod_{i:l_i = 1} N(d_i; 0.08, 0.06^2) }{\sum_{\mathbf{k} \in \{0, 1\}^n} \prod_{i : k_i = 0} N(d_i; 0.02, 0.06^2) \prod_{i:k_i = 1} N(d_i; 0.08, 0.06^2) } \\
&= \frac{ \prod_{i : l_i = 0} \exp\big(-\frac{1}{2\cdot 0.06^2}(d_i - 0.02)^2\big) \prod_{i:l_i = 1} \exp\big(-\frac{1}{2\cdot 0.06^2}(d_i - 0.08)^2\big)}{\sum_{\mathbf{k} \in \{0, 1\}^n} \prod_{i : k_i = 0} \exp\big(-\frac{1}{2\cdot 0.06^2}(d_i - 0.02)^2\big) \prod_{i:k_i = 1} \exp\big(-\frac{1}{2\cdot 0.06^2}(d_i - 0.08)^2\big) }\\
&= \frac{ \exp\Big(-\frac{1}{2\cdot 0.06^2}\big(\sum_{i:l_i = 0}(d_i - 0.02)^2 + \sum_{i:l_i = 1} (d_i - 0.08)^2\big) \Big) }{\sum_{\mathbf{k} \in \{0, 1\}^n} \exp\big(-\frac{1}{2\cdot 0.06^2}\big(\sum_{i:k_i = 0}(d_i - 0.02)^2 + \sum_{i:k_i = 1} (d_i - 0.08)^2\big) \big)}
\end{split}
\end{equation}

We want to simulate from the posterior Mosaic RF $\{l(\mathbf{x});\mathbf{x}\in L_D|\mathbf{d}\}$. Because we have an independent prior, each point in the Mosaic RF will be independent, and we get the following posterior distribution
$$[l_i|d_i]\sim p(l_i|d_i) = \frac{p(d_i|l_i)}{\sum_{l_i'\in L} p(d_i|l_i')} = \begin{cases} \frac{N(d_i;0.02,0.06^2)}{N(d_i;0.02,0.06^2)+N(d_i;0.08,0.06^2)} = 1-p_i, &\text{if } l_i = 0,\\ \frac{N(d_i;0.08,0.06^2)}{N(d_i;0.02,0.06^2)+N(d_i;0.08,0.06^2)} = p_i, &\text{if } l_i = 1\end{cases}; i =1,\dots,n.$$
This means that $[l_i|d_i]$ will be Bernoulli distributed with probability $p_i$ as given above. We can simulate from the posterior distribution by simulating each point from a Bernoulli distribution with given $p_i$. Below, we display 10 realizations of the posterior Mosaic RF.
```{r, echo = F, eval = T, out.width = "20%"}
# Algorithm to sample from model with independent prior
# First, calculate all p_i's based on the observations
p = dnorm(obs, 0.08, 0.06^2)/(dnorm(obs, 0.02, 0.06^2) + dnorm(obs, 0.08, 0.06^2))
# Number of points
n = 75*75

x = rep(1:75,75)
y = rep(1:75,each = 75)
plot_image <- function(l){
  mtrx3d <- data.frame(x = x, y = y, z = ifelse(as.vector(l), "Shale", "Sand"))
  mtrx.melt <- melt(mtrx3d, id.vars = c("x","y"), measure.vars = "z")
  mtrx.melt$value = as.factor(mtrx.melt$value)
  return(ggplot(mtrx.melt, aes(x = x, y = y, fill = value)) +
         geom_tile() + scale_fill_manual("Rock type", values=c("dark blue","dark red")))
}


l = rbinom(n,1,p)
plot_image(l)

l = rbinom(n,1,p)
plot_image(l)

l = rbinom(n,1,p)
plot_image(l)

l = rbinom(n,1,p)
plot_image(l)

l = rbinom(n,1,p)
plot_image(l)

l = rbinom(n,1,p)
plot_image(l)

l = rbinom(n,1,p)
plot_image(l)

l = rbinom(n,1,p)
plot_image(l)

l = rbinom(n,1,p)
plot_image(l)

l = rbinom(n,1,p)
plot_image(l)
```
We see that they are all very similar, and they look reasonable when compared to the observations in figure \ref{fig:obs}. To expand on our analysis, we want to develop expressions for the expectation $E\{\mathbf{l}|\mathbf{d}\}$ and the variances in the diagonal terms of the matrix $\text{Var}\{\mathbf{l}|\mathbf{d}\}$.

We know that each point is independent, and each point conditioned on the observation in that point has a Bernoulli distribution with parameter $p_i$. From theory, we know that if $l_i\sim \text{Bernoulli}(p_i)$, then $E(l_i) = p_i$ and $Var(l_i) = p_i(1-p_i)$. This gives 
$$E\{\mathbf{l}|\mathbf{d}\} = \mathbf{p} = (p_1,\dots,p_n), \quad \text{Var}\{l_i|d_i\} = p_i(1-p_i), i=1,\dots,n.$$

In addition, we want to develop an expression for the maximum marginal posterior predictor $$\text{MMAP}\{\mathbf{l}|\mathbf{d}\} = \hat{\mathbf{l}} = \bigg[\hat{l_i} = \underset{l_i\in\mathbb{L}}{\operatorname{arg max}}\{p(l_i|d_i)\} \bigg]_{i=1,\dots,n}.$$

Because $[l_i|d_i]$ is a Bernoulli distributed variable with parameter $p_i$, we get
$$\hat{l_i} = \begin{cases}0, &\text{if } p_i<0.5,\\ 1, &\text{if } p_i >0.5.\end{cases}$$

Below, we have plotted the expecation, variance and MMAP.
```{r, echo = F, eval = T, out.width = "33%"}
dlist = list(x = 1:75, y = 1:75, z = matrix(p, nrow = 75, ncol = 75))
image.plot(dlist, main = "Expected value of l")

dlist = list(x = 1:75, y = 1:75, z = matrix(p*(1-p), nrow = 75, ncol = 75))
image.plot(dlist, main = "Variance of l")

dlist = list(x = 1:75, y = 1:75, z = matrix(round(p), nrow = 75, ncol = 75))
image.plot(dlist, main = "Maximum marginal posterior predictor")
```

COMMENT ON THE RESULTS

## c)
